---
title: "BIOL668_R_Notebook"
output:
  pdf_document: default
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---
```{r libraries}
#none needed
```


```{r load data}
list.files('~') #are all the files present?
d1=read.table('RTestData.txt', header=TRUE) 
```

```{r functions}
d1 #outputs the whole datafrane
fuso #cant use on its own
d1$fuso  #have to specify column within dataframe
attach(d1) #attaches dataframe to R notebook
fuso #now can use on its own
```

```{r original data}
#observe data
hist(d1$strep) #looks skewed
qqnorm(d1$strep)
qqline(d1$strep, col = "Hotpink", lwd = 2) #not normal
```

```{r log transform data}
#log transformation
lstrep <- log(d1$strep)
qqnorm(lstrep)
qqline(lstrep, col = "Hotpink", lwd = 2) #more normal
#plot histogram
hist(lstrep) #looks more like a normal distribution
```


```{r t-test}
#compare t-tests for orginal vs non-transformed
t.test(d1$strep, d1$fuso) 
t.test(lstrep, d1$fuso)
```
> t.test(d1$strep, d1$fuso) 

	Welch Two Sample t-test

data:  d1$strep and d1$fuso
t = 3.3584, df = 97.994, p-value = 0.001117
alternative hypothesis: true difference in means is not equal to 0
95 percent confidence interval:
 2.116011 8.228433
sample estimates:
mean of x mean of y 
13.958333  8.786111 

> t.test(lstrep, d1$fuso)

	Welch Two Sample t-test

data:  lstrep and d1$fuso
t = -10.219, df = 74.469, p-value = 8.04e-16
alternative hypothesis: true difference in means is not equal to 0
95 percent confidence interval:
 -7.730404 -5.207866
sample estimates:
mean of x mean of y 
 2.316976  8.786111 
 
How did the test statistic and p-value change? Why did the p-value change?
- log transformation reduces variance and the skewness, making any difference appear more significant where p-value p-value = 8.04e-16 for logged vs p-value = 0.001117 for non-transformed.

```{r primer data import}
d2 <- read.csv("primer-1.csv")
attach(d2)
```

```{r inspect data}
d2 #shows the whole dataframe
d2$Plate.Name #shows just the first variable (v1)
```


```{r order}
newdata=d1[order(time),] #orders the column time 
time1=subset(d1, time=='1') #subset the dataframe where time is all equal to 1
```

```{r write files}
write.csv(time1,"time1.csv")

```

# RLab1.2

```{r import r datasets}
attach(airquality)
attach(ToothGrowth)


#observe data
head(airquality)
head(ToothGrowth)

#omit NAs
airquality_df <- na.omit(airquality)

```

# Air quality - Ozone

```{r test for normality Ozone}
hist(airquality_df$Ozone) #not normal
qqnorm(airquality_df$Ozone)
qqline(airquality_df$Ozone, col = "Hotpink", lwd = 2) #more normal
shapiro.test(airquality_df$Ozone) #p-value < 0.05, not normal
```


```{r log ozone transformation airquality}
Ozone_log <- log(airquality_df$Ozone)

hist(Ozone_log) #not normal
qqnorm(Ozone_log)
qqline(Ozone_log, col = "Hotpink", lwd = 2) #more as normal
shapiro.test(Ozone_log) #p-value < 0.05, not normal
```

```{r sqrt ozone transformation airquality}
Ozone_sqrt <- sqrt(airquality_df$Ozone)

hist(Ozone_sqrt) #not normal
qqnorm(Ozone_sqrt)
qqline(Ozone_sqrt, col = "Hotpink", lwd = 2) #more as normal
shapiro.test(Ozone_sqrt) #p-value < 0.05, not normal

```

```{r}
Ozone_cube <- airquality_df$Ozone^(1/3)

hist(Ozone_cube) #looks normal
qqnorm(Ozone_cube)
qqline(Ozone_cube, col = "Hotpink", lwd = 2) #more as normal
shapiro.test(Ozone_cube) #p-value > 0.05, normal
```



# Airquality - Solar.R

```{r test for normality solar}
hist(airquality_df$Solar.R) #not normal
qqnorm(airquality_df$Solar.R)
qqline(airquality_df$Solar.R, col = "Hotpink", lwd = 2) #more normal
shapiro.test(airquality_df$Solar.R) #p-value < 0.05, not normal
```


```{r log solar transformation solar}
solar_log <- log(airquality_df$Solar.R)

hist(solar_log) #not normal, skewed
qqnorm(solar_log)
qqline(solar_log, col = "Hotpink", lwd = 2) #not normal at all
shapiro.test(solar_log) #p-value < 0.05, not normal
```

```{r sqrt solar transformation airquality}
solar_sqrt <- sqrt(airquality_df$Solar.R)

hist(solar_sqrt) #not normal, less skewed
qqnorm(solar_sqrt)
qqline(solar_sqrt, col = "Hotpink", lwd = 2) #not normal
shapiro.test(solar_sqrt) #p-value < 0.05, not normal

```

```{r rank solar transformation airquality}
solar_rank <- rank(airquality_df$Solar.R)

hist(solar_rank) #looks weird
qqnorm(solar_rank)
qqline(solar_rank, col = "Hotpink", lwd = 2) #not normal
shapiro.test(solar_rank) #p-value < 0.05, not normal

```

```{r arcsin solar transformation airquality}
solar_arc <- asin(sqrt(airquality_df$Solar.R)) #NaNs produced??

```

# Airquality - Wind


```{r}
#wind
hist(airquality$Wind) #appears normal
qqnorm(airquality$Wind)
qqline(airquality$Wind, col = "Hotpink", lwd = 2) #not very normal 
shapiro.test(airquality$Wind) #p-value > 0.05, normal
```


# Airquality - Temp
```{r temp}
hist(airquality$Temp) #appears normal
qqnorm(airquality$Temp)
qqline(airquality$Temp, col = "Hotpink", lwd = 2) #normal 
shapiro.test(airquality$Temp) #p-value < 0.05, not normal
```

```{r log temp}
temp_log <- log(airquality_df$Temp)

hist(temp_log) #looks normal-ish
qqnorm(temp_log)
qqline(temp_log, col = "Hotpink", lwd = 2) #appears not normal 
shapiro.test(temp_log) #p-value < 0.05, not normal
```

```{r sqrt temp transformation airquality}
sqrt_temp <- sqrt(airquality_df$Temp)

hist(sqrt_temp) #looks normal
qqnorm(sqrt_temp)
qqline(sqrt_temp, col = "Hotpink", lwd = 2) #normal-ish
shapiro.test(sqrt_temp) #p-value < 0.05,  not normal

```

```{r rank temp transformation airquality}
rank_temp <- sqrt(airquality_df$Temp)

hist(rank_temp) #looks normal
qqnorm(rank_temp)
qqline(rank_temp, col = "Hotpink", lwd = 2) #normal-ish
shapiro.test(rank_temp) #p-value < 0.05,  not normal

```


# Airquality - Month

```{r}
shapiro.test(month) #not numeric, can't apply?
```


```{r}
month_log <- log(airquality_df$Month)

hist(month_log) #not normal
qqnorm(month_log)
qqline(month_log, col = "Hotpink", lwd = 2) #not very normal 
shapiro.test(month_log) #p-value < 0.05, not normal
```

```{r sqrt month transformation airquality}
sqrt_month <- sqrt(airquality_df$Month)

hist(sqrt_month) #looks not normal
qqnorm(sqrt_month)
qqline(sqrt_month, col = "Hotpink", lwd = 2) #normal-ish
shapiro.test(sqrt_month) #p-value < 0.05, not normal

```


```{r rank month transformation airquality}
rank_month <- rank(airquality_df$Month)

hist(rank_month) #looks not normal
qqnorm(rank_month)
qqline(rank_month, col = "Hotpink", lwd = 2) #normal-ish
shapiro.test(rank_month) #p-value < 0.05, not normal

```

# Airquality - Day

```{r}
hist(day) #not normal, skewed
qqnorm(day)
qqline(day, col = "Hotpink", lwd = 2) #not very normal 
shapiro.test(day) #p-value < 0.05, not normal
```


```{r}
day_log <- log(airquality_df$Day)

hist(day_log) #not normal, skewed
qqnorm(day_log)
qqline(day_log, col = "Hotpink", lwd = 2) #not very normal 
shapiro.test(day_log) #p-value < 0.05, not normal
```

```{r sqrt day transformation airquality}
sqrt_day <- sqrt(airquality_df$Day)

hist(sqrt_day) #looks potentially normal
qqnorm(sqrt_day)
qqline(sqrt_day, col = "Hotpink", lwd = 2) #normal-ish
shapiro.test(sqrt_day) #p-value < 0.05, not normal

```

# ANOVA

```{r anova}
#set month as factor
Month <- as.factor(airquality$Month)

#perform anvoa 
anovatemp <- aov(Temp ~ Month, data = airquality) #month is significant
summary(anovatemp)
```

# Airquality pairs
```{r scatterplots and correlations}
#scatterplot of all pairs + shows correlations
pairs(airquality)

```

# perform correlations

```{r specific correlations}
#specified by the assignment
#correlation between ozone and solar
cor.test(airquality$Ozone, airquality$Solar.R) #semi correlated

#correlation between wind and temp
cor.test(airquality$Wind, airquality$Temp) #not correlated


```



```{r modify p-value}
#extracts p-value from the correlations performed above as well
p_values <- c(
  cor.test(airquality$Ozone, airquality$Solar.R)$p.value,
  cor.test(airquality$Wind, airquality$Temp)$p.value
)
p_adjusted_fdr <- p.adjust(p_values, method = "fdr") #fdr
p_adjusted_bonf <- p.adjust(p_values, method = "bonferroni") #bonferroni

p_adjusted_fdr
p_adjusted_bonf

```
output from the above:
> p_adjusted_fdr
[1] 1.793109e-04 5.283194e-09
> p_adjusted_bonf
[1] 3.586217e-04 5.283194e-09



#ToothGrowth

```{r toothgrowth data}
data("ToothGrowth")
head(ToothGrowth)
```

# ToothGrowth - Len

```{r toothgrowth length}
hist(ToothGrowth$len) #looks potentially normal
qqnorm(ToothGrowth$len)
qqline(ToothGrowth$len, col = "Hotpink", lwd = 2) #normal-ish
shapiro.test(ToothGrowth$len) #p-value > 0.05, normal

```

# ToothGrowth - Supp

```{r toothgrowth supp}
ToothGrowth$supp
qqnorm(ToothGrowth$supp) #all NAs?
qqline(ToothGrowth$len, col = "Hotpink", lwd = 2) #not applied
shapiro.test(ToothGrowth$len) #p-value > 0.05, normal?

```

# ToothGrowth - Dose

```{r toothgrowth dose}
hist(ToothGrowth$dose) #looks weird
qqnorm(ToothGrowth$dose)
qqline(ToothGrowth$dose, col = "Hotpink", lwd = 2) #normal-ish
shapiro.test(ToothGrowth$dose) #p-value < 0.05, not normal


#log transform
ToothGrowth_log <- log(ToothGrowth$dose)
hist(ToothGrowth_log ) #looks weird
qqnorm(ToothGrowth_log )
qqline(ToothGrowth_log , col = "Hotpink", lwd = 2) #normal-ish
shapiro.test(ToothGrowth_log ) #still not normal

#sqrt -
ToothGrowth_sqr <- sqrt(ToothGrowth$dose)
hist(ToothGrowth_sqr ) #looks weird
qqnorm(ToothGrowth_sqr )
qqline(ToothGrowth_sqr, col = "Hotpink", lwd = 2) #normal-ish
shapiro.test(ToothGrowth_sqr ) #still not normal

#cube root
ToothGrowth_cube <- (ToothGrowth$dose)^1/3
hist(ToothGrowth_cube ) #looks weird
qqnorm(ToothGrowth_cube )
qqline(ToothGrowth_cube, col = "Hotpink", lwd = 2) #normal-ish
shapiro.test(ToothGrowth_cube ) #still not normal

#rank
ToothGrowth_rank <- rank(ToothGrowth$dose)
hist(ToothGrowth_rank ) #looks weird
qqnorm(ToothGrowth_rank )
qqline(ToothGrowth_rank, col = "Hotpink", lwd = 2) #normal-ish
shapiro.test(ToothGrowth_rank ) #still not normal??????


```

# Anova and Linear regression

```{r anova + linear regression }
#change factor
supp <- as.factor(ToothGrowth$supp)

#ANOVA
anova <- aov(len ~ supp, data = ToothGrowth)

# Display summary
summary(anova) #non significant > 0.05


#linear regression
dose <- as.numeric(ToothGrowth$dose)

# Run Linear Regression
model <- lm(len ~ dose, data = ToothGrowth)
summary(model) #does and len are significant <0.05

```

# toothgrowth pairs
```{r scatterplots and correlations for toothgrowth}
#scatterplot of all pairs + shows correlations
pairs(ToothGrowth) #hard to see? but len and dose appear correlated

```

# perform correlations

```{r specific correlation to see any correlations}
#specified by the assignment
cor.test(ToothGrowth$len, ToothGrowth$dose) #semi correlated



```



```{r modify p-value}
#extracts p-value from the correlations performed above as well
corr_tooth <- cor.test(ToothGrowth$len, ToothGrowth$dose)

#extract p-value
p_value <- corr_tooth$p.value


#adjusts p-value based on method 
p_adjusted_fdr <- p.adjust(p_value, method = "fdr") #fdr
p_adjusted_bonf <- p.adjust(p_value, method = "bonferroni") #bonferroni

#outputs adjusted p-values
p_adjusted_fdr
p_adjusted_bonf

```
