---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: console
---
```{r libraries}
library(ggplot2)
library(vegan)
```



```{r import data}
data <- read.table("RTestData.txt", header = TRUE)
```

# Part A: Univariate statistics with R
Normality check & data transformation

```{r normality of bacteria}
#strep
hist(data$strep) #looks skewed
qqnorm(data$strep)
qqline(data$strep, col = "Hotpink", lwd = 2) #not normal
shapiro.test(data$strep) #p < 0.05, not normal

#prev
hist(data$prev) #looks skewed
qqnorm(data$prev)
qqline(data$prev, col = "Hotpink", lwd = 2) #not normal
shapiro.test(data$prev) #p < 0.05, not normal

#lepto
hist(data$lepto) #looks skewed
qqnorm(data$lepto)
qqline(data$lepto, col = "Hotpink", lwd = 2) #normal-ish
shapiro.test(data$lepto) #p > 0.05, normal

#fuso
hist(data$fuso) #looks skewed
qqnorm(data$fuso)
qqline(data$fuso, col = "Hotpink", lwd = 2) #normal-ish
shapiro.test(data$fuso) #p < 0.05, not normal

#veil
hist(data$veil) #looks skewed
qqnorm(data$veil)
qqline(data$veil, col = "Hotpink", lwd = 2) #not normal
shapiro.test(data$veil) #p < 0.05, not normal

#deepest
hist(data$deepest) #looks normal
qqnorm(data$deepest)
qqline(data$deepest, col = "Hotpink", lwd = 2) #normal
shapiro.test(data$deepest) #p > 0.05, normal
```

```{r data transformation}
#strep
strep_log <- log(data$strep)
hist(strep_log) #looks normal
qqnorm(strep_log)
qqline(strep_log, col = "Hotpink", lwd = 2) #normal
shapiro.test(strep_log) #normal

#prev
prev_log <- log(data$prev + 0.001) #added 0.001 to account for zeros 
shapiro.test(prev_log) #not normal

prev_sqrt <- sqrt(data$prev)
hist(prev_sqrt) #looks normal
qqnorm(prev_sqrt)
qqline(prev_sqrt, col = "Hotpink", lwd = 2) #normal
shapiro.test(prev_sqrt) #normal

#fuso
fuso_log <- log(data$fuso + 0.001) #accounts for zeros
shapiro.test(fuso_log) #not normal

fuso_sqrt <- sqrt(data$fuso)
shapiro.test(fuso_sqrt) #not normal

fuso_rank <- rank(data$fuso)
hist(fuso_rank) #looks normal-ish
qqnorm(fuso_rank)
qqline(fuso_rank, col = "Hotpink", lwd = 2) #normal
shapiro.test(fuso_rank) #normal p > 0.05

#veil
veil_log <- log(data$veil)
shapiro.test(veil_log) #not normal

veil_sqrt <- sqrt(data$veil)
hist(veil_sqrt) #looks normal
qqnorm(veil_sqrt)
qqline(veil_sqrt, col = "Hotpink", lwd = 2) #normal
shapiro.test(veil_sqrt) #normal
``` 


One-way ANOVA and summary and boxplots:
(1) Lepto by time
(2) Strep by time

```{r factor change}
data$time <- factor(data$time)
```


```{r anova}
#anova, summary, boxplot for lepto vs time
anov1 <- aov(lepto ~ time, data = data)
summary(anov1) #time is significant (p < 0.05)
boxplot(lepto ~ time, data = data, main = "Lepto vs Time")

#anova, summary, boxplot for strep_log vs time
anov2 <- aov(strep_log ~ time, data = data)
summary(anov2) #not significant (p > 0.05)
boxplot(strep_log ~ time, data = data, main = "Strep vs Time")
```

Correlation analysis (cor.test)
(1) Strep vs. prev
(2) Strep vs. fuso
(3) Fuso vs. lepto
```{r correlation}
cor.test(strep_log, prev_sqrt) #neg correlated, significant
cor.test(strep_log, fuso_rank) #neg correlated, significant
cor.test(fuso_rank, data$lepto) #neg correlated, not significant
```

Linear regressions
Use the lm command and use plot and abline to produce graphs.
(1) Strep vs. deepest
(2) Fuso vs. deepest

```{r}
#linear regression of strep vs deepest
mod1 <- lm(strep_log ~ deepest, data = data)
plot(mod1) #plots all residuals
abline(mod1, col = "red", lwd = 2)
summary(mod1) #deepest is not significant p > 0.05

#linear regression of fuso vs deepest
mod2 <- lm(fuso_rank ~ deepest, data = data)
plot(mod2)
abline(mod2, col = "red", lwd = 2)
summary(mod2) #deepest is significant p < 0.05
```


# Part B: Use ggplot2 to make pretty graphs
Produce the following graphs:
(1) ggplot histogram of prev variable.
(2) ggplot scatterplot of strep by deepest, colored by time
(3) ggplot scatterplot of fuso by deepest
(4) Same as 3 but use smoothing algorithm

```{r histogram prev}
#fancy histogram of prev
prev_plot <- ggplot(data, aes(x = prev_sqrt)) + geom_histogram(binwidth = 0.5, fill = "white", color = "black") + labs(title = "Prev", x= "Prev") + theme_classic()
prev_plot
```

```{r scatterplot strep vs deepest}
#scatterplot for strep vs deepest color coded by time
plot2 <- ggplot(data, aes(strep_log, deepest, colour = time)) + geom_point() + labs(title = "Strep vs Deepest", x = "Strep", y = "Deepest") + theme_classic()
plot2
```

```{r}
#scatterplot of fuso vs deepest
plot3 <- ggplot(data, aes(fuso_rank, deepest)) + geom_point() + labs(title = "Deepest vs Fuso", x = "Fuso", y = "Deepest") + theme_classic() + geom_smooth(method=lm)
plot3
```



# Part C: Vegan analysis
(1) Calculate alpha diversity (Shannon) for the dataset.
(2) Rarefy the data and make a plot using rarecurve.
(3) Calculate Bray-Curtis dissimilarity matrix, show the matrix and plot a histogram.

```{r new data}
#takes columns of interests and creates new dataframe
newdata = data[2:6]
newdata <- round(newdata)
```

```{r alpha diversity}
shannon <- diversity(newdata) #shannon is the default
shannon #ranges from 1.5 - 3.4, the higher = more diverse 
```

note: rarefaction generates the expected number of species in a small collection of n individuals (or n samples) drawn at random from the large pool of N samples.
```{r rarefraction}
spAbund <- rowSums(newdata) #number of individuals found in each plot 
spAbund

raremin <- min(rowSums(newdata))
raremin #smallest number of observations per sample, 11

#expected rarefied number of species
sRare <- rarefy(newdata, raremin)
sRare

#rarecurve plot
rarecurve(newdata, col = "blue")
```

```{r bray-curtis}
par(mfrow = c(1,2))
bray = vegdist(newdata, "bray") #bray-curtis dissimarility matrix
bray #prints/shows matrix
hist(bray, xlim = range(0,1)) #histogram
```


(1) Make a multidimensional scaling plot using the Bray-Curtis matrix.
(2) Color the plot by status, then time.
```{r}
newnew_data = data[2:8]
```


```{r MDS}
#combines several steps into one
community.mds <- metaMDS(comm = newnew_data, distance = "bray", trace = FALSE, autotransform = FALSE)

plot(community.mds$points)
```

```{r new dataframe}
#extracts x and y and creates a new dataframe
MDS_xy <- data.frame(community.mds$points)

MDS_xy$status <- factor(data$status)
MDS_xy$time <- factor(data$time)
```

```{r new plot}
#mds by status
ggplot(MDS_xy, aes(x = MDS1, y = MDS2, color = status)) +
  geom_point(size = 3) +
  theme_bw() +
  labs(title = "MDS Colored by Status")

#mds by time
ggplot(MDS_xy, aes(x = MDS1, y = MDS2, color = time)) +
  geom_point(size = 3) +
  theme_bw() +
  labs(title = "MDS Colored by Time")
```


